### 2024-10-29 teaching/math240f24/draft.tex
lecture 29 oct

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}
\def\percent{\mathrel{\%}}

\vglue24pt

@@

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}

@@

\proclaim Proposition \advthm. Let $a$ and $b$ be nonzero integers. The set
$$X = \{ s'a + t'b : s',t'\in \ZZ\}$$
is exactly the set of multiples of $d = \gcd(a,b)$.

@@

Let's see a couple of examples. Suppose we want to find out if $9$ is prime or not (we both know it isn't
but play along for a second). In the very first step loop, when $a=2$, we have
$$2^{9-1} = 2^8 = (2^4)^2 = 16^2 \equiv (-2)^2 = 4\pmod 9,$$
so since $4 \not\equiv 1\pmod 9$, we conclude right away that $9$ is not prime.

@@

\proclaim Lemma \advthm. Let $a$ and $b$ be integers.
\medskip
\item{i)} For all positive integers $m$ and $n$,
if $a\equiv b\pmod{mn}$, then $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
\medskip
In the case that $m$ and $n$ are different primes, the converse holds; that is, the following
holds.
\medskip
\item{ii)} For all primes $p$ and $q$ with $p\ne q$, if $a\equiv b\pmod p$
and $a\equiv b\pmod q$, then $a\equiv b\pmod{pq}$.
\medskip

@@

A {\it graph} is a pair $G = (V,E)$ where $V$ is a nonempty set and
$$E \subseteq \bigl\{ \{u,v\} : u,v\in V\bigr\}.$$
The elements of $V$ are called {\it vertices} and the elements of $E$ are
called edges.
If $e = \{u,v\}\in E$, we say that $u$ and $v$ are {\it adjacent}, and
$e$ is {\it incident on} $u$ and $v$. For brevity, we often write $uv$ instead
of $\{u,v\}$.

For instance, we might have $V = \{1,2,3,4\}$ and
$$E = \{12, 13, 14, 24\}.$$
We can draw the graph $G = (V,E)$ as follows:
\midinsert
$$\epsfbox{firstexample.ps}$$
\endinsert
\goodbreak
More examples of graphs are pictured in the figure below.

\midinsert
$$\epsfbox{smallgraphs.ps}$$
\vskip5pt
\caption{Some small graphs.}
\endinsert
\goodbreak

We can define graphs in a more complicated manner. Let $G = (V,E)$
be the graph with $V = \ZZ/5\ZZ$ and $ab\in E$ if and only if
$$a + b \equiv 1\pmod 5\qquad\hbox{or}\qquad a+b \equiv 3\pmod 5.$$
Here is a drawing of $G$:
\midinsert
$$\epsfbox{modfiveexample.ps}$$
\endinsert
\goodbreak

>>>>

\def\datestamp#1#2 {\llap{{\oldstyle #1}{\rm.}{\sc #2}\hskip12pt}}
\def\bar{\overline}
\def\symdiff{\mathrel{\triangle}}
\def\To{\Rightarrow}
\def\Gets{\Leftarrow}
\font\titlefont=cmbx12 at 14 pt
\def\cont#1#2{#1\hskip-2pt\leaders\hbox to 8pt{\hss.\hss}\hfil\hbox to 1em{\hss#2}}
\def\divides{\mid}
\def\percent{\mathrel{\%}}
\def\norm#1{|\!|#1|\!|}
\def\dist{\op{\rm dist}}

@@

\vskip \medskipamount \hbox to\hsize
{\relax \unhcopy \strutbox \cont
{{\tenssbx I. Foundations \kern 3pt}}{2}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 1.\kern .5em Set theory }{3}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 2.\kern .5em Propositional logic}{10}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 3.\kern .5em Predicate logic}{17}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 4.\kern .5em Proofs}{21}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 5.\kern .5em Functions}{28}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 6.\kern .5em Cardinality}{33}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 7.\kern .5em Relations}{37}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx II. Number theory \kern 3pt}}{41}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 8.\kern .5em Division }{42}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 9.\kern .5em Primes }{47}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 10.\kern .5em Modular arithmetic }{51}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 11.\kern .5em Applications of
number theory}{60}}
\medskip
\noindent\hbox to\hsize{\relax \unhcopy \strutbox \cont
{{\tenssbx III. Graph theory \kern 3pt}}{68}}
\smallskip
\hbox to\hsize {\relax \unhcopy \strutbox \cont {\hskip10pt 12.\kern .5em Definitions and
basic notions}{69}}

@@

\edef\propfrogs{\the\thmcount}
\proclaim Proposition \advthm. Let $a$ and $b$ be nonzero integers. The set
$$X = \{ s'a + t'b : s',t'\in \ZZ\}$$
is exactly the set of multiples of $d = \gcd(a,b)$.

@@

\noindent\datestamp{29}{x} \hskip\parindent
Let's see a couple of examples.
Suppose we want to find out if $9$ is prime or not (we both know it isn't
but play along for a second). In the very first step loop, when $a=2$, we have
$$2^{9-1} = 2^8 = (2^4)^2 = 16^2 \equiv (-2)^2 = 4\pmod 9,$$
so since $4 \not\equiv 1\pmod 9$, we conclude right away that $9$ is not prime.

@@

\proclaim Lemma \advthm. Let $a$ and $b$ be integers.
\medskip
\item{i)} For all positive integers $m$ and $n$,
if $a\equiv b\pmod{mn}$, then $a\equiv b\pmod m$ and $a\equiv b\pmod n$.
\medskip\noindent
In the case that $m$ and $n$ are different primes, the converse holds; that is, the following
holds.
\medskip
\item{ii)} For all primes $p$ and $q$ with $p\ne q$, if $a\equiv b\pmod p$
and $a\equiv b\pmod q$, then $a\equiv b\pmod{pq}$.
\medskip

@@

A {\it graph} is a pair $G = (V,E)$ where $V$ is a nonempty set and
$$E \subseteq \bigl\{ \{u,v\} : u,v\in V, u\ne v\bigr\}.$$
The graph $G$ is said to be {\it finite} if and only if $V$ is finite.
The elements of $V$ are called {\it vertices} and the elements of $E$ are
called edges.
If $e = \{u,v\}\in E$, we say that $u$ and $v$ are {\it adjacent}, and
$e$ is {\it incident on} $u$ and $v$. For brevity, we often write $uv$ instead
of $\{u,v\}$. Note that an edge is defined to be a $2$-element set, not an ordered
pair, so $uv = vu$.

Let's start with a small example. Let $V = \{1,2,3,4\}$ and
$$E = \{12, 13, 14, 24\}.$$
We can draw the graph $G = (V,E)$ as follows:
\midinsert
$$\epsfbox{firstexample.ps}$$
\endinsert
\goodbreak
%More examples of graphs are pictured in the figure below.
%\midinsert
%$$\epsfbox{smallgraphs.ps}$$
%\vskip5pt
%\caption{Some small graphs.}
%\endinsert
%\goodbreak

Instead of listing out all the vertices,
we can define graphs by setting conditions for when a pair of vertices should
be connected by an edge.
For instance, let $G = (V,E)$
be the graph with $V = \ZZ/5\ZZ$ and, for $a,b\in V$ with $a\ne b$,
we put $ab\in E$ if and only if
$$a + b \equiv 1\pmod 5\qquad\hbox{or}\qquad a+b \equiv 3\pmod 5.$$
Here is a drawing of $G$:
\midinsert
$$\epsfbox{modfiveexample.ps}$$
\endinsert
\goodbreak
Turns out this graph was simpler than it might have seemed at first glance.
Here is an alternate drawing of $G$:
\midinsert
$$\epsfbox{modfiveexamplesimple.ps}$$
\endinsert
\goodbreak
Note that when we refer to $G$, we are talking
about a set of vertices and edges, not any particular drawing!
There are infinitely many valid drawings of any graph.

Suppose $G = (V,E)$ has a finite number of vertices. Then we can put the elements
of $V$ in bijection with $\{1,\ldots, n\}$. Without loss of generality, we can simply assume
that $\{1,\ldots,n\}$ {\it is} the vertex set of $G$.
The {\it adjacency matrix} of $G$ is then defined to be the symmetric $n\times n$ matrix
obtained by setting
$$A_{ij} = \cases{ 1, & if $ij\in E$,\cr 0, & otherwise.}$$
\endgroup%\headline{\hfil}\footline{\hfil}
For instance, the adjacency matrix of the graph on $\ZZ/5\ZZ$ defined earlier is
$$ A = 
\pmatrix{
0 & 1 & 0 & 0 & 1 \cr
1 & 0 & 0 & 1 & 0 \cr
0 & 0 & 0 & 0 & 1 \cr
0 & 1 & 0 & 0 & 0 \cr
1 & 0 & 1 & 0 & 0 \cr
}.$$

We now define certain important families of graphs. For any integer $n\ge 1$, we define
the {\it complete graph} $K_n$ to be the graph on $n$ vertices with every possible edge present.
Small examples are illustrated in Fig.~2.
\midinsert
$$\epsfbox{kn.ps}$$
\vskip5pt
\caption{Complete graphs $K_n$ for small $n$.}
\endinsert
\goodbreak
\bye
For an integer $n\ge 1$,
we define the {\it Hamming cube} or {\it hypercube} to be the graph $Q_n$
whose vertex set is $\{0,1\}^n$, the set of all binary strings of length $n$,
and whose edge set is the exactly the set of pairs of strings that differ by exactly
one bit. Small examples are illustrated in Fig.~3. (As an exercise, try to figure out
which strings correspond to which vertices!)
\midinsert
$$\epsfbox{qn.ps}$$
\vskip5pt
\caption{Hypercubes $Q_n$ for small $n$.}
\endinsert
\goodbreak

\medskip\boldlabel Degrees and {\mathbold k}-regularity.
The {\it neighbours} of a vertex $v$ are all $u\in V$ such that $uv\in E$.
The {\it degree} of a vertex $v$, denoted by $\deg(v)$, is the number of neighbours
that $v$ has. A graph is said to be {\it $k$-regular} for some $k\in \NN$ if every $v\in V$ has
degree $k$. For example, let $C_n$ denote the graph on $V = \ZZ/n\ZZ$ where
we connect $a\in V$ and $b\in V$ if and only if $a-b\equiv 1\pmod n$
or $a-b\equiv -1\pmod n$. Every vertex $a$ has exactly two neighbours, namely,
the congruence classes of $a-1$ and $a+1$ modulo $n$. So $C_n$ is $2$-regular for all $n$.
In fact, complete graphs and hypercubes are also regular, for in $K_n$, each vertex
is connected to each of the $n-1$ other vertices, and in $Q_n$, each string is
connected to the $n$ other strings obtained by flipping each of its bits.

The following theorem relates vertex degrees to the number of edges.

\edef\thmdegformula{\the\thmcount}
\proclaim Theorem \advthm. Let $G = (V,E)$ be a finite graph. Then
$$\sum_{v\in V} \deg(v) = 2|E|.$$

\proof Consider each edge $e = uv \in E$. It contributes $+2$ to the right-hand side.
On the left-hand side, it contributes $+1$ to the degree $\deg(u)$ of $u$,
and $+1$ to the degree $\deg(v)$ of $v$.\slug

The proof of this theorem is an excellent example of a proof by double counting, where
we have proved an identity by showing that both sides are different ways
of counting the same thing. This theorem has the following corollary.

\parenproclaim Corollary {\advthm} (Handshaking lemma).
In every finite simple graph, the number of vertices having odd degree is even.

\proof We partition $V = V_o \cup V_e$, where $V_o$ comprises all vertices of
odd degree and $V_e$ comprises all vertices of even degree. By the previous theorem,
we have
$$2|E| = \sum_{v\in V} \deg (v) = \sum_{v\in V_o} \deg(v) + \sum_{v\in V_e} \deg(v).$$
Taking this whole identity modulo $2$, we have
$$0 \equiv \sum_{v\in V_o} 1 + \sum_{v\in V_e} 0\pmod 2.$$
Hence
$$|V_o| = \sum_{v\in V_o} 1 \equiv 0\pmod 2,$$
which is what we wanted.\slug

This proposition is often called the handshaking lemma (sometimes this is also used
to refer to the preceding theorem),
since it implies that at any gathering, the number of people who shake hands with
an odd number of people is even. From Theorem~{\thmdegformula} we also able to derive
a corollary that counts the number of edges in $k$-regular graphs.

\proclaim Corollary \advthm. Let $G = (V,E)$ be $k$-regular. Then
$$|E| = {k|V|\over 2}.$$

\proof By Theorem~{\thmdegformula}, we have
$$2|E| = \sum_{v\in V}\deg (v),$$
but $\deg(v) = k$ for all $v\in V$ by $k$-regularity. Hence
$$2|E| = \sum_{v\in V} k = k|V|,$$
and the result follows upon rearranging.\slug

\medskip\boldlabel\datestamp{18}{} Walks, paths, and cycles.
A {\it walk} in $G = (V,E)$ is a sequence $\sigma$ of vertices
$$\sigma = (v_0, v_1, \ldots, v_n)$$
such that $v_i v_{i+1}\in E$ for all $0\le i \le n-1$. The {\it endpoints}
of $\sigma$ are $v_0$ and $v_n$, and the {\it length} of $\sigma$, denoted $|\sigma|$,
is the number of edges traversed, namely, $n$.
The walk is said to be {\it closed} if $v_0 = v_n$ and {\it open} otherwise.
A walk is called a {\it path} if no vertex is repeated.

\proclaim Theorem \advthm. Let $G = (V,E)$ be a graph. If $u$ and $v$ are vertices
such that there exists a walk from $u$ to $v$, then there exists a path from $u$ to $v$.

\proof We perform a minimality argument.
Let $\sigma = (v_0, v_1,\ldots, v_n)$ be a walk from $u$ to $v$ (so
$u = v_0$ and $v = v_n$) of shortest length. We claim that $\sigma$ is in fact a path.
Indeed, suppose for a contradiction that it is not a path; then there is some repeated vertex,
so there exist $i,j\in\{0,1,\ldots,n\}$ such that $i<j$ and $v_i = v_j$. Hence
there is no need to visit any of the vertices between $v_i$ and $v_j$ in $\sigma$,
since $v_i = v_j$ is connected to $v_{j+1}$. Concretely, consider the walk
$$\sigma' = (v_0, v_1,\ldots, v_i, v_{j+1}, \ldots, v_n).$$
Note that $|\sigma'| = |\sigma| - (j-i)$, and $j-i > 0$, so $\sigma'$ is a shorter walk
from $u$ to $v$. But this contradicts our choice of $\sigma$ as a walk from
$u$ to $v$ of shortest length. We conclude that $\sigma$ is a path.\slug

A {\it cycle} is a walk $\sigma = (v_0,v_1,\ldots,v_n)$ of length at least $3$
with $v_0 = v_n$ and no vertex repeated except $v_0 = v_n$.

\edef\propoddwalkoddcycle{\the\thmcount}
\proclaim Proposition \advthm. Let $G = (V,E)$. If $G$ contains a closed walk
of odd length, then it contains a cycle of odd length.

\proof The idea is similar to the proof of the previous theorem. We perform a
minimality argument. Let
$\sigma = (v_0, v_1,\ldots,v_n)$ be an odd-length closed walk in $G$, and choose
this walk to have minimal odd length (i.e, any shorter walk has even length). We shall
prove that $\sigma$ is a cycle.

For a contradiction, suppose that $\sigma$ is not a cycle, so that
there exist indices $i$ and $j$ with $i<j$ such that $v_i = v_j$.
Consider the two closed walks
$$\sigma_1 = (v_0, v_1, \ldots, v_i, v_{j+1}, \ldots, v_n)$$
and
$$\sigma_2 = (v_i, v_{i+1}, \ldots, v_j).$$
Both are shorter than $\sigma$, so by the minimality of $\sigma$,
$|\sigma_1|$ and $|\sigma_2|$ are even. But this implies that
$|\sigma| = |\sigma_1|+|\sigma_2|$ is even. This contradicts the hypothesis that $|\sigma|$
is odd.

We conclude that $\sigma$ is a cycle.\slug

\medskip\boldlabel Connectedness.
We say a graph $G = (V,E)$ is {\it connected} if for all $u,v\in V$ there exists
a walk (and hence a path) from $u$ to $v$. All of the graphs we have
drawn so far have been connected. Here is a graph that is disconnected:
\midinsert
$$\epsfbox{disconnected.ps}$$
\endinsert
\noindent This picture is meant to depict one graph on $8$ vertices, not three
different graphs.

The small examples of $K_n$ and $Q_n$ in Figs.~2 and~3 were all connected,
so it is natural to assume that $K_n$ and $Q_n$ are connected for all $n$.
This is indeed the case.

\proclaim Proposition \advthm. For all $n\ge 1$, the graphs $K_n$ and $Q_n$
are connected.

\proof Any vertices $u$ and $v$ in the vertex set of $K_n$ belong to the
edge $uv$, so we have the path $(u, v)$ of length $1$ between $u$ and $v$.
This shows that $K_n$ is connected.

Now let $u$ and $v$ be any two vertices in $K_n$. Suppose there are $m$
bits that differ between $u$ and $v$. Then we can flip them one by one to
change $u$ into $v$. This represents a path of length $m$ from $u$ to $v$, since there
is an edge in $Q_n$ between any two strings in $Q_n$ that differ at exactly one
bit.\slug

Let $G = (V,E)$ and
let $u,v\in V$. We say that {$v$ is reachable from $u$} if there exists a walk
from $u$ to $v$. If we define a relation $\sim$ on $V$ by setting
$u\sim v$ if and only if $v$ is reachable from $u$, then $\sim$ is an equivalence
relation (prove this as an exercise!) and the equivalence
classes are called the {\it connected components} of $G$.

Here is an example. Let $G = (\ZZ, E)$, where for $i,j\in \ZZ$ with $i<j$, we have
$ij\in E$ if and only if $j-i \in \{9,15\}$. That is, every edge either
hops forwards or backwards in the number line in increments of $9$ or $15$.
Starting at $n\in \ZZ$, we can reach any vertex $m$ that is of the form
$$ m = n + 15s + 9t$$
for some $s,t\in \ZZ$. By Proposition~{\propfrogs} (this was the proposition
about frogs and lilypads, if that jogs your memory), the integers
representable
as $15s+9t$ for some $s,t\in \ZZ$ are exactly the multiples of
$\gcd(9,15) = 3$. So in fact, from $n$ one can reach any integer $m$
of the form $n+3l$ for some $l\in \ZZ$; that is, one can reach any integer
$m$ with $m\equiv n\pmod 3$. So the three connected components in this
graph are $[0]_3$, $[1]_3$, and $[2]_3$, the equivalence classes of integers
modulo $3$.

\advsect Triangles and bipartite graphs

A {\it subgraph} of $G = (V,E)$ is a graph $G' = (V',E')$ such that
$V'\subseteq V$ and $E'\subseteq E$ where for all $e = uv\in E'$, we have
$u,v\in V'$. For example, the {\it triangle graph} $C_3$ is a subgraph of $K_4$ (draw it!).
In fact, any graph on $n$ vertices or less is a subgraph of $K_n$, since
$K_n$ contains all possible edges among $n$ vertices.

We now consider an extremal question concerning
subgraphs. An extremal question is one of the form, {\sl What is the
extremal (maximum or minimum) number of objects we can have, subject to some restriction?}
The extremal question we ask is, {\sl What is the maximum number of edges that a graph
on $n$ vertices can have, assuming that it does not contain a triangle as a subgraph?}
To answer this question, we need the following inequality, which is one of the
most important inequalities in all of mathematics. It is named for A.~Cauchy,
who in 1821 proved the version we shall record below,
and H.~Schwarz, who proved a version for integrals in 1888.

\parenproclaim Theorem {\advthm} (Cauchy--Schwarz inequality). For all
$u_1,\ldots,u_n, v_1,\ldots,v_n\in \RR$, we have
$$\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
\le \biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr).$$

In the proof below, you will need to recall from your high school days that
for a quadratic polynomial $ax^2 + bx + c$, the {\it discriminant} is
defined to be the quantity $b^2 - 4ac$. (It appears under the square-root sign
in the quadratic formula.) If it is negative, the polynomial has no real roots,
if it is positive, the polynomial has two real roots, and if it is zero, the polynomial
has one real root.

\proof If $u_i = 0$ for all $1\le i\le n$, then both sides are $0$ and the inequality
is trivially true. So suppose that at least one of the $u_i$ is nonzero. Then
$$p(x) = (u_1 x + n_1)^2 + (u_2 x + v_2)^2 + \cdots + (u_n x+v_n)^2$$
is a quadratic polynomial in the variable $x$. Rewriting $p(x)$ in summation notation,
we have
$$\eqalign{
p(x) &= \sum_{i=1}^n (u_ix + v_i)^2\cr
&= \sum_{i=1}^n \bigl( u_i^2 x^2 + 2u_i v_i + v_i^2\bigr)\cr
 &= \biggl(\sum_{i=1}^n u_i^2 \biggr) x^2 +
2\biggl(\sum_{i=1}^n u_i v_i\biggr) x + \biggl( \sum_{i=1}^n {v_i}^2\biggr).\cr
}$$
Since it is defined as a sum of squares, $p(x)\ge 0$ for all $x\in \RR$, so it has
at most one real root.
This means its discriminant is nonpositive; in other words,
$$4\biggl(\sum_{i=1}^n u_i v_i\biggr)^2
- 4\biggl( \sum_{i=1}^n {u_i}^2\biggr) \biggl( \sum_{i=1}^n {v_i}^2\biggr) \le 0.$$
Dividing through by $4$ and then rearranging gives exactly the inequality we sought.\slug

The Cauchy--Schwarz inequality can also be proved using some results you might
know from linear algebra.
Defining vectors $u,v\in \RR^n$ by
$u = (u_1,\ldots,u_n)$ and $v = (v_1,\ldots,v_n)$, the left-hand side of the
Cauchy--Schwarz inequality is
$(u\cdot v)^2$ and the right-hand side is $\norm{u}^2\norm{v}^2$. But,
letting $\theta$ is the angle between $u$ and $v$,
then $u\cdot v = \cos\theta \norm{u}\norm{v}$, and applying the fact that
$|\cos\theta|\le 1$, we easily deduce the desired inequality.

We are now able to derive the maximum number of edges in a triangle-free graph on
$n$ vertices. The following theorem was proved by W.~Mantel in 1907.

\parenproclaim Theorem {\advthm}. (Mantel's theorem). Let $G = (V,E)$ be a graph
not containing a triangle as a subgraph. Then
$$|E| \le \biggl\lfloor {|V|^2\over 4}\biggr\rfloor.$$

\proof Consider the sum
$$\sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr).$$
The term $\deg(u)$ appears in the sum exactly once for every edge incident on $u$;
that is, it appears $\deg(u)$ times. This is true for all $u\in V$, so
we conclude that
$$\sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr) = \sum_{u\in V} \deg(u)^2.$$
On the other hand, since $G$ contains no triangle, for every pair of vertices $u$
and $v$, the set of neighbours of $u$ is disjoint from the set of neighbours of $v$.
So $\deg(u)+\deg(v)\le |V|$, and we have
$$\sum_{u\in V} \deg(u)^2 = \sum_{uv\in E} \bigl(\deg(u)+\deg(v)\bigr)\le |V|\cdot |E|.$$
By Theorem~{\thmdegformula} and the Cauchy--Schwarz inequality in that order,
we have
$$\eqalign{
\bigl(2|E|\bigr)^2 &= \Bigl(\sum_{u\in V} \deg(u)\Bigr)^2 \cr
&= \Bigl(\sum_{u\in V} \deg(u)\cdot 1\Bigr)^2 \cr
&\le \Bigl(\sum_{u\in V} \deg(u)^2\Bigr) \Bigl(\sum_{u\in V} 1^2\Bigr) \cr
&= |V| \Bigl(\sum_{u\in V} \deg(u)^2\Bigr).\cr
}$$
Hence
$$4|E|^2 \le |V| \Bigl(\sum_{u\in V} \deg(u)^2\Bigr) \le |V|^2\cdot |E|.$$
Rearranging this gives us
$$|E|\le {|V|^2\over 4},$$
and we can take the floor function on the right-hand side, since $|E|$ must be an
integer.\slug

So if $G = (V,E)$
has $|E|\ge \lfloor |V|^2/4\rfloor$, there must be a triangle subgraph in $G$.
Now let $|V|$ be even, so that $|V|^2/4$ is an integer. Does there exist a graph
$G$ with exactly $|V|^2/4$ edges such that $G$ does not contain a triangle
as a subgraph? This brings us to the definition of a bipartite graph.

\medskip\boldlabel\datestamp{19}{} Bipartite graphs.
A graph $G = (V,E)$ is called {\it bipartite} if there is a partition
$V = A\cup B$ of the vertex set (so $A\cap B = \emptyset$) called
the {\it bipartition} such that each edge has one endpoint in $A$ and one endpoint
in $B$. For example, hypercubes are bipartite.

\proclaim Proposition \advthm. For all $n\ge 1$, the hypercube $Q_n$ is bipartite.

\proof Let $Q_n = (V,E)$. Every element $s\in V$ corresponds to a string
$S = (s_1, \ldots, s_n)$ of bits, where $s_i\in \{0,1\}$ for all $1\le i\le n$.
Define
$$V_0 = \bigl\{ s\in V : s_1 + \cdots + s_n \equiv 0 \pmod 2\bigr\}$$
and
$$V_1 = \bigl\{ s\in V : s_1 + \cdots + s_n \equiv 1 \pmod 2\bigr\}.$$
It is clear that $V_0 \cup V_1 = V$ and $V_0 \cap V_1 = \emptyset$, so
this is a bipartition of the vertex set, and for every $e = s_1s_2\in E$,
the strings $s_1$ and $s_2$ differ in exactly one bit, so if $s_1\in V_0$,
then $s_2\in V_1$, and vice versa.\slug

The {\it complete bipartite graph} $K_{m,n}$ is a graph with $V = V_m \cup V_n$, where
$|V_m|=m$ and $|V_n| = n$, and $E$ is the set $\{uv : u\in V_m, v\in V_n\}$ of
all possible edges between $V_m$ and $V_n$. Fig.~4 illustrates two small examples
of complete bipartite graphs.

\midinsert
$$\epsfbox{kmn.ps}$$
\vskip5pt
\caption{The complete bipartite graphs $K_{2,3}$ and $K_{4,3}$.}
\endinsert
\goodbreak
When $n$ is even, the graph $K_{n/2,n/2}$ has $n$ vertices and exactly $n^2/4$ edges.
When $n$ is odd, the graph $K_{(n+1)/2, (n-1)/2}$ has $n$ vertices and
$$ {n+1\over 2} \cdot {n-1\over 2} = {n^2-1\over 4} = {n^2\over 4}-{1\over 4}
= \biggl\lfloor {n^2\over 4} \biggr\rfloor$$
edges. (The last equality here follows from the fact that any odd $n$ is congruent to either $1$ or
$3$ modulo $4$, which means its square is congruent to $1$ modulo $4$.)

The calculation above shows that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$
always has $n$ vertices and $\lfloor n^2/4\rfloor$ edges. It is the largest possible number
of edges such that Mantel's theorem does not apply; that is, from Mantel's theorem alone,
we do not know whether
there must be a triangle in $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, but adding
a single edge to $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$, Mantel's theorem tells
us there must be a triangle in the resulting graph.

We will actually be able to prove that $K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not
contain any triangles. This means that the lower bound of $\lfloor n^2/4\rfloor$ in Mantel's theorem
is the best possible one.

First we need a lemma.

\proclaim Lemma \advthm. A graph $G=(V,E)$ is bipartite if and only if all of its connected components
are bipartite.

\proof Let $G_1 = (V_1, E_1), G_2 = (V_2,E_2),\ldots, G_n = (V_k, E_k)$ be the
connected components of $G$.

If each connected component is bipartite, say,
$V_i = A_i\cup B_i$ is a bipartition of $V_i$ for all $1\le i\le n$, then
$A = \bigcup_{i=1}^n A_i$ and $B = \bigcup_{i=1}^n B_i$ is a bipartition of $V$,
and every $e\in E$ has one endpoint in $A$ and one endpoint in $B$.

Now suppose that there is some component $G_i$ that is not bipartite.
Now let $V = A\cup B$
be any partition of $V$ into disjoint nonempty sets. Then $A_i = A\cap V_i$
and $B_i = B\cap V_i$ form a partition of $V_i$
into two nonempty sets. But $G_i$ is not bipartite, so there is some $e\in E$
that either has both endpoints in $A_i$ or both endpoints in $B_i$. This means
that $G$ is not bipartite either.\slug

As a matter of fact, we shall prove something much stronger than just the fact that
$K_{\lceil n/2 \rceil, \lfloor n/2 \rfloor}$ does not contain any triangles.
The following theorem characterises the family of bipartite graphs as {\it exactly} those graphs not
containing odd-length cycles.
Its proof shall require the following metric imposed on a graph $G$. For all $u,v\in V$
define the {\it distance} $\dist(u,v)$ between $u$ and $v$ to be the length of the shortest path
from $u$ to $v$. If no such path between $u$ and $v$ exists, we set $\dist(u,v) = \infty$.

\proclaim Theorem \advthm. A graph is bipartite if and only if it does not
contain any cycles of odd length.

\proof First, assume that $G=(V,E)$ is bipartite; let $V = A\cup B$ be the bipartition
of the vertex set. Let $\sigma$ be a cycle in $G$. Each edge changes side (either
$A$ to $B$ or $B$ to $A$), so in order for the starting and ending vertices of this
cycle to be the same, $|\sigma|$ must be even.

Now assume that $G = (V,E)$ has no odd-length cycles. To show that $G$ is bipartite,
it suffices to show that each of its connected components is bipartite (by the
previous lemma), so without
loss of generality we may assume that $G$ is connected.
This means that $\dist(u,v)< 0$ for all $u,v\in V$.

Select one vertex $h\in V$ and set
$$ V_0 = \{v\in V : \dist(h,v)\equiv 0\pmod 2\}$$
and
$$ V_1 = \{v\in V : \dist(h,v)\equiv 1\pmod 2\}.$$
Let $e = uv\in E$.
Consider a closed walk $\sigma$ that follows a shortest path from $u$ to $h$, then a shortest
path from $h$ to $v$, and finally the edge $e$ going from $v$ to $u$. We have
$$|\sigma| = \dist(u,h) + \dist(h,v) + 1.$$
But since $G$ contains no odd-length cycles, by (the contrapositive of)
Proposition~{\propoddwalkoddcycle}, any closed walk in $G$ must have even length.
This means that $\dist(u,h)$ and $\dist(h,v)$ are not the same modulo $2$; that is,
either $u\in V_0$ and $v\in V_1$ or vice versa.\slug

\advsect Trees

A graph $G$ is called a {\it forest} if it has no cycles. A connected forest is called
a {\it tree}. A vertex of degree $1$ in a forest is called a {\it leaf}. Fig.~5 contains
an example of a tree.

\midinsert
$$\epsfbox{spidery.ps}$$
\vskip5pt
\caption{A tree with $25$ leaves (count them!).}
\endinsert
\goodbreak
Our first proposition
concerning trees is that every tree containing at least one edge has at least two leaves.

\edef\proptwoleaves{\the\thmcount}
\proclaim Proposition \advthm. Let $F$ be a forest with at least one edge (and hence at least
two vertices). There are at least two leaves in $F$.

\proof Let $\sigma = (v_0,v_1,\ldots,v_n)$ be a path of maximum length in $F$.
Since there is at least one edge in $F$, we have $|\sigma|\ge 1$. The claim is that
$\deg(v_0) = \deg(v_n) = 1$. Indeed, for a contradiction, assume that one of these
two endpoints has degree at least two. Without loss of generality, suppose it is
$v_0$ (we can reverse the path if needed to make this the case). So there exists
$u\in V$ such that $uv_0$ is an edge. If $u = v_i$ for some $1\le i\le n$, then
there is a cycle $(v_i, v_0, v_1,\ldots, v_i)$, contradicting the fact that $F$ is a forest.
So for all $1\le i\le n$, $u\ne v_i$. This means that
$(u,v_0, \ldots, v_n)$ is a strictly longer path than $\sigma$, contradicting the maximality
of $\sigma$.\slug

The following gives a characterisation of trees.

\proclaim Proposition \advthm. A graph $G = (V,E)$ is a tree if and only if
for all $u,v\in V$, there exists a unique path from $u$ to $v$ in $G$.
